
--- 
title: "Brianna's Exploratory Analysis Script"
author: "Brianna Correa"
date: "04/06/2025"
format: html
--- 

## Introduction

In this report, we answer three research questions related to the Palmer Penguin dataset. The dataset includes measurements of penguins from the Palmer Station, Antarctica, focusing on body mass, flipper length, sex, and species.

## Research Questions:
1. Does sex predict body mass when controlling for species?
2. Is there a relationship between body mass and flipper length?
3. Is there a relationship between body mass and flipper length?

Hopefully by the end of project 3, I can also ask another question: 

4. Do foraging patterns differ by species or sex?


## Documented workflow

In this project it made more sense to separate cleaning our datasets and exploring them in an analysis file later. This was due to real errors we had to fix and accomdate for that would have prevented us from analyzing it to begin with. In an effort to carefully document this, notes have been left in several readmes related to these files as well as pretty specific notes within the script that explains what each command does.

By cutting the cleaning and our analysis separately I was able to create scripts that were more logicially organized and less confusing. I have created three scripts iwthin this exploratory analysis folder that is meant to help explore and answer the questions related to the data as discussed in the manuscript. 

I have kept the raw data but am working with the clean data, so that we can track the changes made between them but also keep a documentation of what data was originally collected in case those corrections were unnecessary.  

# Analysis 

## Setup

### This chunk loads the necessary libraries and your data:

```{r}
# Load necessary libraries
library(here)
library(ggplot2)
library(dplyr)
library(car)

# Load data
data_path <- here("Data", "Processed_data", "processeddata.rds")
dat <- readRDS(data_path)

# Preview the dataset
head(dat)

# Using the here() Function
# In this project, I ended up having to learn how to use the here() function to make file paths more portable and ensure that my analysis could be easily reproduced across different environments.

# Prior to using here(), I struggled with setting absolute file paths that worked consistently across different machines. The here() function allows the construction of file paths that are relative to the root directory of your project, avoiding hard-coded paths that could break when moving or sharing the project via a repo.

# The here() function works by starting from the root directory of your project, which is especially useful when organizing files in subdirectories (e.g., in the Data, Code, Results folders). Instead of specifying the full path like C:/Users/yourname/..., I can use:

data_path <- here("Data", "Processed_data", "processeddata.rds")
dat <- readRDS(data_path)


```{r}

# Penguin Analysis Script: Question 1

# Create a boxplot to visualize flipper length by species
ggplot(dat, aes(x = Species, y = `Flipper Length (mm)`, fill = Species)) + 
  geom_boxplot() +
  labs(title = "Flipper Length by Species", 
       x = "Species", 
       y = "Flipper Length (mm)") +
  theme_minimal()

# Save the plot
ggsave(filename = here("Results", "flipper_length_species_boxplot.png"))

# Perform ANOVA
lm_fit <- lm(`Flipper Length (mm)` ~ Species, data = dat)
anova_result <- anova(lm_fit)
print(anova_result)

# Save results
saveRDS(anova_result, file = here("Results", "flipper_length_species_anova.rds"))

# Check residuals
residuals <- lm_fit$residuals
ggplot(data.frame(residuals), aes(x = residuals)) +
  geom_histogram(binwidth = 2, fill = "pink", color = "black", alpha = 0.7) +
  labs(title = "Residuals Histogram", x = "Residuals", y = "Frequency") +
  theme_minimal()

ggsave(filename = here("Results", "residuals_histogram.png"))

# Levene's Test
leveneTest(`Flipper Length (mm)` ~ Species, data = dat)

message("The ANOVA results indicate whether there is a statistically significant difference in flipper length between species. 
If the p-value is less than 0.05, it suggests that species differ significantly in their flipper lengths.")

print("Script has run successfully! :)")
```

```{r}

# Penguin Analysis Script: Question 2


# Linear regression
lm_fit <- lm(`Body Mass (g)` ~ Sex + Species, data = dat)

# Summary
summary(lm_fit)

# Boxplot
ggplot(dat, aes(x = Sex, y = `Body Mass (g)`, fill = Sex)) +
  geom_boxplot() +
  labs(title = "Body Mass by Sex", x = "Sex", y = "Body Mass (g)") +
  theme_minimal()

# Save plot
ggsave(filename = here("Results", "body_mass_sex_boxplot.png"))

# Save model
saveRDS(lm_fit, file = here("Results", "body_mass_sex_model.rds"))

# Levene’s test
levene_test_result <- leveneTest(`Body Mass (g)` ~ Sex * Species, data = dat)
print(levene_test_result)

message("The regression results indicate whether sex predicts body mass when controlling for species. 
If the p-value is less than 0.05, it suggests that sex significantly predicts body mass.")

print("Script has run successfully! :)")
```

```{r}

# Penguin Analysis Script: Question 3


# Clean NA values
dat_clean <- dat %>%
  filter(!is.na(`Body Mass (g)`) & !is.na(`Flipper Length (mm)`))

# Linear regression
lm_fit <- lm(`Body Mass (g)` ~ `Flipper Length (mm)`, data = dat_clean)

# Summary
summary(lm_fit)

# Scatterplot with regression line
ggplot(dat_clean, aes(x = `Flipper Length (mm)`, y = `Body Mass (g)`)) +
  geom_point(aes(color = Species), size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Relationship between Body Mass and Flipper Length",
       x = "Flipper Length (mm)", y = "Body Mass (g)") +
  theme_minimal()

# Save plot
ggsave(filename = here("Results", "body_mass_vs_flipper_length.png"))

# Save model
saveRDS(lm_fit, file = here("Results", "body_mass_vs_flipper_length_model.rds"))

message("The regression results show the relationship between body mass and flipper length. 
If the p-value is less than 0.05, it suggests that there is a statistically significant relationship between the two variables.")

print("Script has run successfully! :)")

```














