---
title: "Brianna's Analysis Script"
author: "Brianna Correa"
date: "03/03/2025"
output: html_document
--- 

# Documented workflow

In this project it made more sense to separate cleaning our datasets and exploring them in an analysis file later. This was due to real errors we had to fix and accomdate for that would have prevented us from analyzing it to begin with. In an effort to carefully document this, notes have been left in several readmes related to these files as well as pretty specific notes within the script that explains what each command does.

By cutting the cleaning and our analysis separately I was able to create scripts that were more logicially organized and less confusing. I have created three scripts iwthin this exploratory analysis folder that is meant to help explore and answer the questions related to the data as discussed in the manuscript. 

I have kept the raw data but am working with the clean data, so that we can track the changes made between them but also keep a documentation of what data was originally collected in case those corrections were unnecessary.  

# Analysis 

## Setup

This needs to run to load the R script so we can include the code chunks below.

```{r }
#| cache = FALSE
knitr::read_chunk('statistical_analysis.R')
```

Load needed packages and filepaths.  

```{r setup}
#| message: FALSE
#| warning: FALSE
```

Load custom functions.

```{r functions}
```

Load the data.

```{r loaddata}
```






