---
title: "Brianna Correa: Cleaning Script"
author: "Brianna Correa"
date: "2023-03-03"
format: html
---

## Processing script (data cleaning)

### Setup

Load needed packages. Make sure they are installed.

```{r}
require(dplyr)   # for data processing/cleaning
require(tidyr)   # for data processing/cleaning
require(skimr)   # for nice visualization of data 
```
<!--
"---" has a special meaning in quarto, notice that it is part of the YAML header. Your quarto will not run properly with these sprinkled throughout the markdown. 
You must successfully run at the command line to produce the .html version:
quarto render processingfile_v1.qmd 
--- 
-->

### Data loading

Start R in the same directory that this script resides in: `project_root/Code/Processing_code`.
If needed, check your working directory with `getwd()`.

**Never** put a `setwd()` inside your code — it's bad etiquette and will likely break for others.

```{r}
# Path to data using relative paths
data_location <- "../../Data/Raw_data/penguin_raw_dirty.csv"
data_path <- "../../Data/Raw_data/"
```
<!--
---
-->

### Load data

Use `check.names=FALSE` to preserve original column names (which include spaces/parentheses).

```{r}
rawdata <- read.csv(data_location, check.names=FALSE)
```

<!--
---
-->

### Check data

Use the data dictionary for context.

```{r}
dictionary <- read.csv(paste(data_path, "datadictionary.csv", sep=""))
print(dictionary)
```

Basic overviews:

```{r}
dplyr::glimpse(rawdata)
summary(rawdata)
head(rawdata)
skimr::skim(rawdata)
```
<!--
---
-->

### Cleaning

#### Species names

There are 9 unique species — we expect only 3. Let's examine and clean them.

```{r}
unique(rawdata$Species)
d1 <- rawdata
```

Fix mispellings:

```{r}
ii <- grep("PengTin", d1$Species)
d1$Species[ii] <- "Adelie Penguin (Pygoscelis adeliae)"
d1$Species <- sub("gTin", "guin", d1$Species)
unique(d1$Species)
```

Shorten species to common names:

```{r}
d1$Species <- dplyr::case_when(
  grepl("Adelie", d1$Species) ~ "Adelie",
  grepl("Chinstrap", d1$Species) ~ "Chinstrap",
  grepl("Gentoo", d1$Species) ~ "Gentoo",
  TRUE ~ NA_character_
)
```
<!--
---
-->
#### Continuous data

Fix `"missing"` in `Culmen Length (mm)` and convert to numeric:

```{r}
cl <- d1$`Culmen Length (mm)`
cl[cl == "missing"] <- NA
cl <- as.numeric(cl)
d1$`Culmen Length (mm)` <- cl
```

Check data:

```{r}
skimr::skim(d1)
hist(d1$`Culmen Length (mm)`)
plot(d1$`Body Mass (g)`, d1$`Culmen Length (mm)`)
```

Fix large `Culmen Length (mm)` values (likely off by 10):

```{r}
d2 <- d1
cl <- d2$`Culmen Length (mm)`
cl[!is.na(cl) & cl > 300] <- cl[!is.na(cl) & cl > 300] / 10
d2$`Culmen Length (mm)` <- cl
```

Re-check:

```{r}
skimr::skim(d2)
hist(d2$`Culmen Length (mm)`)
plot(d2$`Body Mass (g)`, d2$`Culmen Length (mm)`)
```

Fix tiny mass values:

```{r}
d3 <- d2
mm <- d3$`Body Mass (g)`
mm[mm < 100] <- NA
nas <- which(is.na(mm))
d3 <- d3[-nas, ]
```

Check again:

```{r}
skimr::skim(d3)
hist(d3$`Body Mass (g)`)
plot(d3$`Body Mass (g)`, d3$`Culmen Length (mm)`)
```

Convert some variables to factors:

```{r}
d3$Species <- as.factor(d3$Species)
d3$Sex <- as.factor(d3$Sex)
d3$Island <- as.factor(d3$Island)
skimr::skim(d3)
```
<!--
---
-->
### Final data check

Make histograms/density plots for continuous data by discrete groups to check for strange categories or value ranges.
<!--
---
-->
### Save cleaned data

```{r}
processeddata <- d3

save_data_location <- "../../Data/Processed_data/processeddata.rds"
saveRDS(processeddata, file = save_data_location)

save_data_location_csv <- "../../Data/Processed_data/processeddata.csv"
write.csv(processeddata, file = save_data_location_csv, row.names=FALSE)
```
<!--
---
-->
### Notes

* Dealing with NA or "bad" data: removing faulty/missing rows is one approach. You could alternatively keep partially missing individuals depending on your analysis goals.
